---
title: "1.2.3: Working With Text"
subtitle: "Fall Into Bioinformatics, 2019"
author: "Steve Pederson"
date: "16 April 2019"
output:
  ioslides_presentation: 
    css: ../custom.css
    fig_caption: yes
    logo: ../images/UoA_logo_col_vert.png
    widescreen: yes
---

```{r, loadPackages, echo = FALSE, include = FALSE}
library(knitr)
opts_chunk$set(echo = TRUE, include = TRUE, 
               warning = FALSE, message = FALSE, 
               out.width = 800, fig.align = "center",
               results = 'hide')
```


# Text Manipulation

## Text Manipulation

- The package `stringr` contains functions for text manipulation 
    + Another core package from `tidyverse`
- Some basic knowledge of regular expressions is helpful
- Functions `str_detect()`, `str_extract()`, `str_replace()`
- Alternatives to `grepl()`, `grep()`, `gsub()` etc from `base`

## `stringr::str_detect()`

```{r}
library(tidyverse)
x <- c("Hi Mum", "Hi Mother")
```

- `str_detect()` returns a logical vector

```{r}
str_detect(string = x, pattern = "Mum")
str_detect(string = x, pattern = "Hi")
```

## `stringr::str_detect()`

We can use common `regex` syntax:

- Alternatives are specified with `[]`

```{r}
str_detect(x, "h")
str_detect(x, "[Hh]")
```

- Wild-cards are specified using `.`

```{r}
str_detect(x, "Mo")
str_detect(x, "M.")
```

## `stringr::str_extract()`

We can use `str_extract()` to extract patterns

```{r}
str_extract(string = x, pattern = "Hi M.")
```

This can be helpful if no matches are found

```{r}
str_extract(x, "Mum")
```


## `stringr::str_replace()`

Common syntax for extracting/modifying text strings

```{r}
str_replace(x, pattern = "Mum", replacement = "Dad")
```

1. Searching the `string` "Hi Mum" for the `pattern` "Mum", and
2. Replacing the first instance of "Mum" with "Dad"

## `stringr::str_replace()`

We can specify wild-cards as `.`

```{r}
str_replace(x, "M.", "Da")
```

We can also match any number of wild-cards by using `+`

```{r}
str_replace(x, "M.+", "Dad")
```

## `stringr::str_replace()`

We can also **capture** words/phrases/patterns using `(pattern)`

```{r}
str_replace(x, "(Hi) (M.+)", "\\2! \\1!")
```

Patterns are numbered in the order they are "captured"

## `stringr::str_replace()`

We can also specify alternatives instead of wild-cards (`[]`)

```{r}
str_replace(x, "[Mm]", "b")
```

- `str_replace()` only replaces the first match in a string
- `str_replace_all()` replaces **all** matches

```{r}
str_replace_all(x, "[Mm]", "b")
```

## `stringr::str_replace()`

Alternative patterns can be specified using the conventional `OR` symbol `|`

```{r}
str_replace(x, "(Mum|Mother)", "Maternal Parent")
```

## More Helpful Functions

```{r}
str_count(x, "[Mm]")
str_length(x)
str_split_fixed(x, pattern = " ", n = 2)
str_to_lower(x)
str_to_title("a bad example")
str_pad(c("1", "10", "100"), width = 3, pad = "0")
```

# Factors

## Factors

A common data type in statistics is a *categorical variable* (i.e. a `factor`)

- Data will be a set of common groups

```{r}
pet_vec <- c("Dog", "Dog", "Cat", "Dog", "Cat")
```

- This is a `character` vector

## Factors

- We can simply coerce this to a vector of factors
- Categories will automatically be assigned alphabetically

```{r}
pet_factors <- as.factor(pet_vec)
pet_factors
```

We can manually set these categories as `levels`

```{r}
pet_factors <- factor(pet_vec, levels = c("Dog", "Cat"))
```

## Factors

- These are actually stored as integers
- Each integer corresponds to a `level`

```{r, results='hide'}
str(pet_factors)
as.integer(pet_factors)
as.character(pet_factors)
```

## Factors

**What would happen if we think a `factor` is a `character`, and we use it to select values from a `vector`/`matrix`/`data.frame`?**

## Factors

**What would happen if we think a `factor` is a `character`, and we use it to select values from a `vector`/`matrix`/`data.frame`?**

```{r}
names(pet_vec) <- pet_vec
pet_vec
pet_vec[pet_factors]
pet_vec[as.character(pet_factors)]
```

*This is why I'm very cautious about `read.csv()` and the standard `data.frame` etc*
