---
title: "1.2 Loading Difficult Data"
subtitle: "Fall Into Bioinformatics, 2019"
author: "Steve Pederson"
date: "15 April 2019"
output:
  ioslides_presentation: 
    css: ../custom.css
    fig_caption: yes
    logo: ../images/UoA_logo_col_vert.png
    widescreen: yes
---

```{r, loadPackages, echo = FALSE, include = FALSE}
library(knitr)
opts_chunk$set(echo = TRUE, include = TRUE, 
               warning = FALSE, message = FALSE, 
               out.width = 800, fig.align = "center")
```

# Loading Difficult Data

## Setup

If you've started a new session since last time:

```{r}
library(tidyverse)
```


## What if the data we have isn't nice?

- Missing values might be given a value (e.g. 9999, "NA")
- Column names might be missing
- File may have comments
- Irrelevant columns
- May be structural issues in the file (Different column numbers)
- White-space in cells

## What if the data we have isn't nice?

- Make sure the file `transport.csv` is in your `data` folder
- Navigate to the file in your `Files` pane
- Click on the file and choose `View File`

**What problems do we face here?**

```{r, echo=FALSE}
opts_chunk$set(eval = FALSE)
```

## What if the data we have isn't nice? 

**What problems do we face here?**

1. A comment in the first line
2. No column names
3. Missing data encoded as "-" in the 4^th^ line
4. A redundant column

## What if the data we have isn't nice? 

Let's try writing code for this instead of using the GUI

**This will fail**

```{r}
transport <- read_csv("data/transport.csv")
```

- `R` uses the first row to guess how many columns there are
    - The comment is indicating 1 column
    
## Removing Comments

- We can tell `R` to ignore any lines beginning with `#`
- Set the comment argument using `comment = "#"`

```{r}
transport <- read_csv("data/transport.csv", comment = "#")
transport
```

 Now `R` is guessing the correct number of columns $\implies$ the file will load

**What does all that red (or blue) stuff mean?**

## Data without column names

- `R` has assumed the first row contains column names
- We can tell `R` to ignore these using: `col_names = FALSE`

```{r}
transport <- read_csv("data/transport.csv", 
                      comment = "#", 
                      col_names = FALSE)
transport
```

**What has `R` used for column names?** 

## Missing Data

**What impact has the missing data in `X5` had?**

- We can correctly assign missing data as `NA` (`na = "-"`)

```{r}
transport <- read_csv("data/transport.csv", 
                      comment = "#", 
                      col_names = FALSE, 
                      na = "-")
transport
```

## Ignoring Columns

```{r, echo=FALSE}
opts_chunk$set(eval = TRUE)
```

- We can also use the code `-` to skip a column
- We can leave `R` to guess any remaining columns using `?`
- Use the `col_types` argument: `col_types = "-?????"`

```{r, results='hide'}
transport <- read_csv("data/transport.csv",
                      comment = "#", 
                      col_names = FALSE, 
                      na = "-", 
                      col_types = "-?????")
transport
```

## Specifying Columns

- Or we can specify the exact type of data in each column
- Numeric columns can be specified as `n`
- Text or `character` columns can be specified as `c`

```{r}
transport <- read_csv("data/transport.csv", 
                      comment = "#", 
                      col_names = FALSE, 
                      na = "-", 
                      col_types = "-ccnnc")
transport
```

## Specifying Columns

- Let's accidentally **get the final column wrong** (use `n` instead of `c`)

```{r,eval=FALSE}
transport <- read_csv("data/transport.csv",
                      comment = "#", 
                      col_names = FALSE, 
                      na = "-",
                      col_types = "-ccnnn")
transport
```

**NB: No warning will be given if a numeric column contains non-numeric characters**


## Specifying Columns

Let's change that back to the correct code:

```{r}
transport <- read_csv("data/transport.csv", 
                      comment = "#", 
                      col_names = FALSE, 
                      na = "-", 
                      col_types = "-ccnnc")
transport
```

## Setting Column Names

- My fix is to supply a `vector` of names

```{r}
myNames <- c("X1", "gender", "name", "weight", "height", "method")
transport <- read_csv("data/transport.csv", 
                      comment = "#", 
                      col_names = myNames, 
                      na = "-", 
                      col_types = "-ccnnc")
transport
```


## The `c()` function

The most common function in `R` is `c()`

- This stands for `combine`
- Combines all values into a single `R` object, or `vector`
- If left empty, it is equivalent to `NULL`

```{r, results='markup'}
c()
```

## The `c()` function

- I have used this to create a vector of column names 
- Used the assignment operator `<-` to assign this vector

**What would happen if I gave too many or too few names?**

- We need to be careful here...

## Exporting Data

After we've edited a file, we might also wish to export it

```{r, eval=FALSE}
?write_csv
```

- This is a wrapper for `write_delim()`
- Can export `.csv`, `.txt`, `.tsv` etc.
- Individual `R` objects can be exported using `write_rds()`

## Exporting Data

The best way to export this is:

```{r}
write_csv(transport, "data/transport_clean.csv")
```

## Bonus Challenge

Download the file `geneCounts.out` (output from featureCounts)

1. Try to import this using the GUI
2. Try to import by using `read_delim()`
    + Set this to skip the `Chr`, `Start`, `End` and `Strand` Columns
    + Try to modify the column names using the function `basename()`
    + Remove `_Fem_hisat2_sorted.bam` from the end of the column names

<div class="footer" style="text-align:center;width:25%">
[Home](http://uofabioinformaticshub.github.io/Intro_R_FIB_2019)
</div>