---
title: "1.3 Plots For Publication"
subtitle: "Fall Into Bioinformatics, 2019"
author: "Steve Pederson"
date: "15 April 2019"
output:
  ioslides_presentation: 
    css: ../custom.css
    fig_caption: yes
    logo: ../images/UoA_logo_col_vert.png
    widescreen: yes
---

```{r, loadPackages, echo = FALSE, include = FALSE}
library(knitr)
library(tidyverse)
opts_chunk$set(echo = TRUE, include = TRUE, 
               warning = FALSE, message = FALSE, fig.align = "center",
               results = 'hide', fig.show = 'hide')
transport <- read_csv("../data/transport_clean.csv") %>%
  filter(!is.na(height))
```

# Making Professional Plots

## Making Professional Plots

Two main points:

1. Tweaking the look of the previous plots
2. Plot Resolution and Font Sizes

## The package `ggplot2` | themes

`ggplot2` uses themes to control the overall appearance

```{r}
ggplot(transport, aes(x = gender, y = height, fill = gender)) +
  geom_boxplot() +
  theme_bw()
```

My default is `theme_bw()`

- Removes the background grey, prints labels in black etc.

## The package `ggplot2` | themes

```{r}
ggplot(transport, aes(x = gender, y = height, fill = gender)) +
  geom_boxplot() +
  theme_classic()
```

```{r}
ggplot(transport, aes(x = gender, y = height, fill = gender)) +
  geom_boxplot() +
  theme_void()
```

## Setting a default theme

You can set a default theme for a workspace or session

```{r}
theme_set(theme_bw())
```

Now all plots in the workspace will use `theme_bw()`

## The package `ggplot2` | themes

The `theme()` function is also where you set:

- `axis.text`, `legend` attributes etc.
- Often uses _elements_ to set an attribute

```{r}
?theme
```

## The package `ggplot2` | themes

Changing text within themes uses `element_text()`

```{r}
?element_text
```

```{r}
ggplot(transport, aes(x = gender, y = height, fill = gender)) +
  geom_boxplot() +
  theme_bw() +
  theme(axis.text.x = element_text(
      angle = 90, hjust = 1, vjust = 0.5
  ))
```

## The package `ggplot2` | themes

Changing backgrounds and outlines uses `element_rect()`

```{r}
ggplot(transport, aes(x = gender, y = height, fill = gender)) +
  geom_boxplot() +
  theme_bw() +
  theme(legend.background = element_rect(
      fill = "yellow", colour = "black"
  ))
```

## The package `ggplot2` | themes

To remove all attributes use `element_blank()` in place of `element_rect()`, `element_text()` or `element_line()`

```{r}
ggplot(transport, aes(x = gender, y = height, fill = gender)) +
  geom_boxplot() +
  theme_bw() +
  theme(panel.grid = element_blank())
```

## Legends

We can move the legend to multiple places:

```{r}
ggplot(transport, aes(x = gender, y = height, fill = gender)) +
  geom_boxplot() +
  theme_bw() +
  theme(legend.position = "bottom")
```

## Legends

Or we can use co-ordinates to place it inside the plotting region

```{r}
ggplot(transport, aes(x = gender, y = height, fill = gender)) +
  geom_boxplot() +
  theme_bw() +
  theme(legend.position = c(0.85, 0.15))
```

The plotting region is assumed to have width and height of 1

## Other plot attributes | Axes

We can also edit axes, fills, outlines etc. using `scale_...()` layers

```{r}
ggplot(transport, aes(x = gender, y = height, fill = gender)) +
  geom_boxplot() +
  theme_bw() +
  scale_y_log10(limits = c(100, 200), 
                breaks = c(100, 125, 150, 175)) 
```

## Other plot attributes | Axes

We can turn off or modify plot expansion

```{r}
ggplot(transport, aes(x = gender, y = height, fill = gender)) +
  geom_boxplot() +
  theme_bw() +
  scale_y_continuous(expand = c(0, 0)) 
```

- The first `a = 0` multiplies the scale by `1 + a`
- The second `b = 0` adds $\pm$`b` to the axis extrema

## Other plot attributes | Fill and Outline Colours

```{r}
ggplot(transport, aes(x = gender, y = height, fill = gender)) +
  geom_boxplot() +
  theme_bw() +
  scale_fill_manual(values = c("green", "blue"))
```

Colours can also be specified using hexadecimal codes

```{r, include=TRUE,results='asis'}
rgb(1, 0, 0)
```

# Saving Images

## Exporting Figures | Using `ggsave()`

The main image formats are `jpeg`, `png` and `tiff`

- `R` can also export `svg` and `pdf`
- `ggplot2` has the function `ggsave()`

```{r}
?ggsave
```

The `Plots` Tab is the default graphics device

## Exporting Figures | Using `ggsave()`

```{r}
ggplot(transport, aes(x = gender, y = height, fill = gender)) +
  geom_boxplot() +
  theme_bw() +
  scale_fill_manual(values = c("green", "blue"))
ggsave("HeightByGender.png")
```

- `ggsave` defaults to 300dpi 
- Defaults to the size of the `Plots` Tab

## Exporting Figures| Using `ggsave()`

- Change the size manually by setting the `width` and `height` attributes

```{r}
ggsave("HeightByGender.png", width = 18, height = 18,
       units = "cm")
```

- Getting the font size right can take ages
- Need to set correctly when we make the plot

## Exporting Figures | Writing Directly

- We initiate another graphics device using `png()`, `jpeg()`, `pdf()` etc.
- Turn the device off after creating the image using `dev.off()`
- Nothing will appear in the `Plots` tab

## Exporting Figures | Writing Directly

```{r}
png("HeightByGender.png", width = 18, height = 18, 
    units = "cm", res = 300)
ggplot(transport, aes(x = gender, y = height, fill = gender)) +
  geom_boxplot() +
  theme_bw() +
  theme(text = element_text(size = 16)) +
  scale_fill_manual(values = c("green", "blue"))
dev.off()
```

## Exporting Figures | Writing Directly

Try to decide how big the plot will be in your final document

- Export using that size
- `pdf` and `svg` output are *vector*-based not *pixel*-based
- Scale very well with size

# Exercises

## Exercises

- Try loading and exploring the additional files:
- `toothData`, `PCR` and `RealTimeData`


---

<div class="footer" style="text-align:center;width:25%">
[Home](http://uofabioinformaticshub.github.io/Intro_R_FIB_2019/)
</div>